State-space analysis of a continuous gravitational wave source with a pulsar timing array: inclusion of the pulsar terms

### 2024-01-24 manuscript/manuscript_2/paper.tex
actual version passed to AM

We define a root-mean-square relative error to quantify the accuracy of the one-dimensional posteriors with respect to the injection value, viz.
\begin{equation}
\Delta_{\rm M}(\theta) = \sqrt{\frac{1}{K}\sum_{k=1}^{K} \frac{ \left[\theta_{k,\rm M} - \theta_{\rm inj}\right]^2}{\theta_{\rm inj}}} \ . \label{eq:mean_rel_error}
\end{equation}
In Equation \eqref{eq:mean_rel_error} $\theta_{k,\rm M}$ is the $k$-th sample of the one-dimensional posterior for parameter $\theta$ returned by nested sampling, and the subscript $\rm M \in \{ \text{Earth, psr \& Earth} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively. The total number of samples is $K$ and $\theta_{\rm inj}$ is the true injection value. The error $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_compare_high} is summarised in Table \ref{tab:posterior_errors}. We find that the estimates are more accurate using the pulsar terms, i.e. $\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for $\theta$ in $\boldsymbol{\theta}_{\rm gw}$. In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models, with $\Delta_{\rm Earth}(\Omega)-\Delta_{\rm psr}(\Omega) = 6.7 \times 10^{-3}$. The improvement from including the pulsar terms is largest for $\iota$, with $\Delta_{\rm Earth}(\iota) - \Delta_{\rm psr \& Earth}(\iota)  = 0.24$. Whilst we present only a single noise realisation in this section, the improvements from including the pulsar terms are generally of a comparable magnitude across different realisations. \newline

@@

Figure \ref{fig:bayes1} reveals an approximate quadratic relationship $\beta \propto h_0^2$ for $h_0 \gtrsim 10^{-14}$ for both $\mathcal{M}_{\rm Earth}$ and $\mathcal{M}_{\rm psr \& Earth}$. Moreover, we obtain $\beta_{\rm psr \& Earth} > \beta_{\rm Earth}$ for all $h_0$. That is, for a given $h_0$, $\mathcal{M}_{\rm psr \& Earth}$ provides greater evidence for a GW signal in the noisy data than $\mathcal{M}_{\rm Earth}$. The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 3.2 \times 10^{-15}$ for  $\mathcal{M}_{\rm Earth}$ and $h_0 \gtrsim 2.8 \times 10^{-15}$ for  $\mathcal{M}_{\rm psr \& Earth}$, a relative improvement in the minimal detectable strain of $14\%$. The minimum detectable strain and the relative improvement through using $\mathcal{M}_{\rm psr \& Earth}$ are particular to the system in Table \ref{tab:parameters_and_priors} and the realisation of $\boldsymbol{Y}$. They are influenced in general by $T_{\rm obs}$, $T_{\rm cad}$, ${\boldsymbol{\theta}}_{\rm gw}$, and ${\boldsymbol{\theta}}_{\rm psr}$. A full parameter sweep is postponed to future work, after the analysis scheme in this paper is upgraded from ingesting pulse frequencies to pulse TOAs to enable a like-for-like comparison with standard PTA analyses. In the interim, we note that a $13\%$ improvement in sensitivity when including the pulsar terms is comparable to the improvements of $5\%$ achieved by \cite{Zhupulsarterms}.

>>>>

We define a root-mean-square relative error to quantify the accuracy of the one-dimensional posteriors with respect to the injection value, viz.
\begin{equation}
\Delta_{\rm M}(\theta) = \sqrt{\frac{1}{K}\sum_{k=1}^{K} \frac{ \left[\theta_{k,\rm M} - \theta_{\rm inj}\right]^2}{\theta_{\rm inj}}} \ . \label{eq:mean_rel_error}
\end{equation}
In Equation \eqref{eq:mean_rel_error} $\theta_{k,\rm M}$ is the $k$-th sample of the one-dimensional posterior for parameter $\theta$ returned by nested sampling, and the subscript $\rm M \in \{ \text{Earth, psr\&Earth} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively. The total number of samples is $K$ and $\theta_{\rm inj}$ is the true injection value. The error $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_compare_high} is summarised in Table \ref{tab:posterior_errors}. We find that the estimates are more accurate using the pulsar terms, i.e. $\Delta_{\rm Earth}(\theta) > \Delta_{\rm psr \& Earth}(\theta)$ for $\theta$ in $\boldsymbol{\theta}_{\rm gw}$. In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models, with $\Delta_{\rm Earth}(\Omega)-\Delta_{\rm psr \& Earth}(\Omega) = 6.7 \times 10^{-3}$. The improvement from including the pulsar terms is largest for $\iota$, with $\Delta_{\rm Earth}(\iota) - \Delta_{\rm psr \& Earth}(\iota)  = 0.24$. Whilst we present only a single noise realisation in this section, the improvements from including the pulsar terms are generally of a comparable magnitude across different realisations. \newline

@@

Figure \ref{fig:bayes1} reveals an approximate quadratic relationship $\beta \propto h_0^2$ for $h_0 \gtrsim 10^{-14}$ for both $\mathcal{M}_{\rm Earth}$ and $\mathcal{M}_{\rm psr \& Earth}$. Moreover, we obtain $\beta_{\rm psr \& Earth} > \beta_{\rm Earth}$ for all $h_0$. That is, for a given $h_0$, $\mathcal{M}_{\rm psr \& Earth}$ provides greater evidence for a GW signal in the noisy data than $\mathcal{M}_{\rm Earth}$. The GW source is detectable with decisive evidence ($\beta \geq 10$) for $h_0 \gtrsim 3.2 \times 10^{-15}$ for  $\mathcal{M}_{\rm Earth}$ and $h_0 \gtrsim 2.8 \times 10^{-15}$ for  $\mathcal{M}_{\rm psr \& Earth}$, a relative improvement in the minimal detectable strain of $14\%$. The minimum detectable strain and the relative improvement through using $\mathcal{M}_{\rm psr \& Earth}$ are particular to the system in Table \ref{tab:parameters_and_priors} and the realisation of $\boldsymbol{Y}$. They are influenced in general by $T_{\rm obs}$, $T_{\rm cad}$, ${\boldsymbol{\theta}}_{\rm gw}$, and ${\boldsymbol{\theta}}_{\rm psr}$. A full parameter sweep is postponed to future work, after the analysis scheme in this paper is upgraded from ingesting pulse frequencies to pulse TOAs to enable a like-for-like comparison with standard PTA analyses. In the interim, we note that a $14\%$ improvement in sensitivity when including the pulsar terms is comparable to the improvements of $5\%$ achieved by \cite{Zhupulsarterms}.
